Low Complexity Adaptive Beamformer\\ for Active Sonar Imaging

### 2015-08-30 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:08:30	14:11:20	(Sun30.Aug)
la2pdf commit - 2015:08:30	14:41:21	(Sun30.Aug)

\ifTODO
   \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
   \setlength\marginparsep{1pt}
   \setlength\marginparwidth{35pt}
   \newlength\marginparwidthsmall
   \setlength\marginparwidthsmall{\marginparwidth}
   \addtolength\marginparwidthsmall{-7pt}
   \newcommand\todo[1]{%
      \addtocounter{todoidx}{1}%
      {\color{Red}\bf(\thetodoidx{})}%%\fbox{\bf\thetodoidx{}}}%
      \marginpar{%
         {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\scriptsize #1}}}}

@@

@@

MVDR seeks to minimize the power of the beamformer, under the constraint of a unity gain in the desired direction $\phi$~\cite{Capon1969}:
%
\begin{align}
\underset{\w[n]}{\argmin}\, E\{|z[n]|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
\text{subject to } \w[n]\a_\phi &= 1.\label{eq_mvdr_definition}
\end{align}
%
Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x\x\H\}$ is the spatial covariance matrix. This is a convex optimization problem with the solution:
%
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi},\label{weights}
\end{gather}
where $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. The problem lies in estimating and inverting this spatial covariance matrix. We have described the exact steps in~\cite{Buskenes2014}. To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}. For true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}. Finally, to improve robustness to parameter errors we apply $d$ percent diagonal loading~\cite{Cox1987,Maksym1979}. The steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.

@@

\begin{figure*}[t]\centering%
\includegraphics[width=\linewidth]{gfx/mvdr_selected_windows_holmengraa.pdf}%
\caption{\emph{Determining LCA window type}. Typical frequency responses for windows computed by a robust MVDR.\newline
\emph{Left:}\hfill
\parbox[t]{.95\linewidth}{MVDR sectorscan image of the oiltanker Holmengraa, using data collected from the 32 element HISAS1030 sonar attached to the HUGIN AUV. MVDR was run with $K=1$ temporal oversampling, $L=16$ subarray size, $d=1\%$ diagonal loading and 8 times lateral oversampling.}\protect\\\hspace{\textwidth}
\emph{Right:}\hfill
\parbox[t]{.95\linewidth}{MVDR window responses computed from 40x40 pixel groups in the shadow, highlight and speckle region of the Holmengraa image. The responses are overlayed each other. The dashed red lines at -13dB and -43dB marks the peak sidelobe levels of an unsteered rectangular and Hamming window, respectively. Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3dB in highlight. The phase varies mostly in highlight.} }\label{mvdr_selected_windows}
\end{figure*}

@@

\begin{figure*}[t]%\centering%
\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds.pdf}%
\caption{Window parameter boundaries. LCA images using an oversampled image and a relatively large window database with $N_\beta=20$ window types, each steered $N_\phi=19$. }\label{oversampling_mosaic_bounds}
\end{figure*}

@@

A much less complex alternative to MVDR is the low complexity adaptive (LCA) beamformer. It iterates through a set of $P$ windows and selects the window $p$ that best fulfills the minimum variance criteron:
%
\begin{align}
\underset{p}{\argmin}\, E\{|z_p[n]|^2\} = \underset{p}{\argmin}\, E\big\{|\w_{p}\H\x[n]|^2\big\}\nn
\qquad\text{subject to}\qquad \w\H\a_\phi = 1.\label{eq_lca_definition}
\end{align}\label{lca_criterion}
%
Note how closely LCA is related to the MVDR definition in \eq{eq_mvdr_definition}. The optimization criterion and constraint is the same, but LCA has a finite and discrete solutionspace for the weights. As will be demonstrated in section \ref{lca_window_function}, LCA will perform similarly to MVDR because a robustified MVDR implementation seems to be constrained to window functions similar to the ones we let LCA choose from.

In practice, we estimate the beamformer output power as
%
\begin{align*}
E\big\{|z_p[n]|^2\big\} = \frac{1}{N_k} \sumb{n'=n-K}{n+K} | z_p[n'] |^2
\end{align*}
%
which computes the sample power average over $N_k=2K+1$ temporal samples. This is the same as we did for MVDR. Our signal is only a few samples long after matched filtering so we have set $K=1$ for both MVDR and LCA.

If extra flexibility is desired, we can also estimate the variance by a weighted combination of local pixels
%
\begin{align}
E\big\{|z_p[x,n]|^2\big\} = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} w_\text{prf}[x',n']\big| z_p[x',n'] \big|^2\end{align}
%
where $N_x = 2X+1$ is the number of azimuth lines to average over, and $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function. \todo{Might want to be careful with this when we steer the windows, or drop it altogether}

\begin{itemize}
\item \emph{Window function}
\item \emph{Parameter boundaries}
\item \emph{Number of windows needed}
\item \emph{Oversampling need}
\item \emph{Image quality}
\end{itemize}

@@

To allow LCA some added flexibility we also create variations of each window that are steered a little to each side.
%
\begin{align}
\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\\
\end{align}
%
where we define the steering vector $\a_\phi$ as:
%
\begin{align}
\a_\phi = \bmat{
1 \\
e^{-j\frac{md}{\lambda}\sin(\phi)} \\
\vdots\\
e^{-j\frac{(M-1)d}{\lambda}\sin(\phi)}
}.
\end{align}
%
Here $d$ is the element spacing, $\lambda$ is the wavelength and $\phi$ is the steering amount. The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain. It ensures unit gain in the direction of interest as required by (\ref{eq_lca_definition}). Since the signal-to-noise ratio is constant for a specific window (same $\beta$-value), this normalization also proportionally increases the incoherent noise gain.

@@

We call the upper bound for steering $\phi_{\text{-3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window. Using a numerical search for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values, we obtained a third degree polynomial fit to these points:
%
\begin{align}
\phi_\text{-3dB}(\beta) \approx -0.002894\beta^3 + 0.02451\beta^2 + -0.01308\beta + 0.2980
\end{align}
%
Henceforth any steering will be specified relative to this upper bound. With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.

@@

While MVDR is free to choose non-symmetric windows, we can observe from \Fig{mvdr_selected_windows} that it prefers symmetric ones. In shadow and speckle the windows are fully symmetric, while in highlight the windows are approximately symmetric within the opening angle of the sonar.

@@

$\phi$ controls the amount of steering.

The Kaiser window parameter $\beta$ adjusts the compromise between resolution and sidelobe suppression. At $\beta=0$ it yields a rectangular window, and at and $\phi$ controls the amount of steering.. $\beta=0$ ) and a gaussian curve (

>>>>

\ifTODO
   \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
   \setlength\marginparsep{1pt}
   \setlength\marginparwidth{35pt}
   \newlength\marginparwidthsmall
   \setlength\marginparwidthsmall{\marginparwidth}
   \addtolength\marginparwidthsmall{-7pt}
   \newcommand\todo[1]{%
      \addtocounter{todoidx}{1}%
      {\color{Red}\bf(\thetodoidx{})}%%\fbox{\bf\thetodoidx{}}}%
      \marginpar{%
         {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\raggedright\scriptsize #1}}}}

@@

\newcommand\minus{\scalebox{0.75}[1.0]{$-$}}

@@

MVDR seeks to minimize the power of the beamformer, under the constraint of a unity gain in the desired direction $\phi$~\cite{Capon1969}:
%
\begin{align}
\underset{\w[n]}{\argmin}\, E\{\big|z[n]\big|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\R[n]\w\H[n]\nn
\text{subject to } \w[n]\a_\phi &= 1.\label{eq_mvdr_definition}
\end{align}
%
Here $\a_\phi$ is a steering vector, $E\{\cdot\}$ is the expectation operator, and $\R[n] = E\{\x\x\H\}$ is the spatial covariance matrix. This is a convex optimization problem with the solution:
%
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a_\phi}{\a_\phi\T\Ri[n]\a_\phi},\label{weights}
\end{gather}
where $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. The problem lies in estimating and inverting this spatial covariance matrix. We have described the exact steps in~\cite{Buskenes2014}. To avoid signal cancellation we apply spatial averaging by computing a mean covariance matrix from a set of subarrays with length $L$~\cite{Kailath1985}. For true speckle statistics we perform temporal averaging over $N_k = 2K+1$ temporal samples~\cite{Synnevag2009a}. Finally, to improve robustness to parameter errors we apply $d$ percent diagonal loading~\cite{Cox1987,Maksym1979}. The steps are also needed to ensure that the covariance matrix is numerically well conditioned and hence invertible.

@@

\begin{figure*}[t]\centering%
\includegraphics[width=\linewidth]{gfx/mvdr_selected_windows_holmengraa.pdf}%
\caption{\emph{Determining LCA window type}: Typical frequency responses for windows computed by a robust MVDR.\newline
\emph{Left:}\hfill
\parbox[t]{.95\linewidth}{MVDR sectorscan image of the oiltanker Holmengraa, using data collected from the 32 element HISAS1030 sonar attached to the HUGIN AUV. MVDR was run with $K=1$ temporal oversampling, $L=16$ subarray size, $d=1\%$ diagonal loading and 8 times lateral oversampling.}\protect\\\hspace{\textwidth}
\emph{Right:}\hfill
\parbox[t]{.95\linewidth}{MVDR window responses computed from 40x40 pixel groups in the shadow, highlight and speckle region of the Holmengraa image. The responses are overlayed each other. The dashed red lines at -13dB and -43dB marks the peak sidelobe levels of an unsteered rectangular and Hamming window, respectively. Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3dB in highlight. The phase varies mostly in highlight.} }\label{mvdr_selected_windows}
\end{figure*}

@@

\begin{figure*}[t]%\centering%
\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds.pdf}%
\caption{\emph{Determining Kaiser parameter boundaries}: LCA images created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied. The upper left image is equal to DAS with rectangular window. While \Fig{mvdr_selected_windows} suggests that an upper bound 
\newline
\emph{Upper bound $\beta$:}\hfill
\parbox[t]{.9\linewidth}{.}\newline
\emph{Right:}\hfill
\parbox[t]{.95\linewidth}{.}
}\label{oversampling_mosaic_bounds}
\end{figure*}

@@

A much less complex alternative to MVDR is the low complexity adaptive (LCA) beamformer. It iterates through a set of $P$ windows and selects the window $p$ that best fulfills the minimum variance criteron:
%
\begin{align}
\underset{p}{\argmin}\, E\{\big|z_p[n]\big|^2\} = \underset{p}{\argmin}\, E\big\{\big|\w_{p}\H\x[n]\big|^2\big\}\nn
\qquad\text{subject to}\qquad \w\H\a_\phi = 1.\label{eq_lca_definition}
\end{align}\label{lca_criterion}
%
Note how closely LCA is related to the MVDR definition in \eq{eq_mvdr_definition}. The optimization criterion and constraint is the same, but LCA has a finite and discrete solutionspace for the weights. As will be demonstrated in section \ref{lca_window_function}, LCA will perform similarly to MVDR because a robustified MVDR implementation seems to be constrained to window functions similar to the ones we let LCA choose from.

In practice, we estimate the beamformer output power as
%
\begin{align*}
E\big\{\big|z[n]\big|^2\big\} = \frac{1}{N_k} \sumb{n'=n-K}{n+K} \big| z[n'] \big|^2
\end{align*}
%
which computes the sample power average over $N_k=2K+1$ temporal samples. This is the same as we did for MVDR. Our signal is only a few samples long after matched filtering so we have set $K=1$ for both MVDR and LCA.

If extra flexibility is desired, we can also estimate the variance by a weighted combination of local pixels
%
\begin{align}
E\big\{\big|z[x,n]\big|^2\big\} = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} w[x',n']\big| z[x',n'] \big|^2
\end{align}
%
where $N_x = 2X+1$ is the number of azimuth lines to average over, and $w_\text{prf}[x',n']$ is a normalized 2 dimensional weight function. \todo{Might want to be careful with this when we steer the windows, or drop it altogether}
% 
% \begin{itemize}
% \item \emph{Window function}
% \item \emph{Parameter boundaries}
% \item \emph{Number of windows needed}
% \item \emph{Oversampling need}
% \item \emph{Image quality}
% \end{itemize}

@@

To allow LCA some added flexibility we also create variations of each window that are steered a little to each side.
%
\begin{align}
\w_{\beta,\phi} = \frac{\f_\beta\H\diag{(\a_\phi)}}{\f_\beta\H\a_\phi}\\
\end{align}
%
where diag($\a_\phi$) is a diagonal matrix constructed from the steering vector $\a_\phi$:
%
\begin{align}
\a_\phi = \bmat{
1 \\
e^{-j\frac{md}{\lambda}\sin(\phi)} \\
\vdots\\
e^{-j\frac{(M-1)d}{\lambda}\sin(\phi)}
}.
\end{align}
%
Here $d$ is the element spacing, $\lambda$ is the wavelength and $\phi$ is the steering amount. The normalization factor $\f_\beta\H\a_\phi$ is the reciprocal of the window's coherent gain. It ensures unit gain in the direction of interest as required by (\ref{eq_lca_definition}). Since the signal-to-noise ratio is constant for a specific window (same $\beta$-value), this normalization also proportionally increases the incoherent noise gain.

@@

We call the upper bound for steering $\phi_{\text{-3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window. Using a numerical search for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values, we obtained a third degree polynomial fit to these points:
%
\begin{align}
\phi_\text{3dB}(\beta) \approx \minus{}0.002894\beta^3 + 0.02451\beta^2 - 0.01308\beta + 0.2980
\end{align}
%
Henceforth any steering will be specified relative to this upper bound. With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.

@@

While MVDR is free to choose non-symmetric windows, we can observe from \Fig{mvdr_selected_windows} that it prefers symmetric ones. In shadow and speckle the windows are fully symmetric, while in highlight the windows are approximately symmetric within the opening angle of the sonar. The peak sidelobe levels are between that of the rectangular and Hamming window.

This correspond to Kaiser windows with $\beta$-values of 0 and 2, respectively.
\Fig{oversampling_mosaic_bounds_beta} illustrates which $\beta$-value LCA prefers for different regions in the cross image.

@@
